Gradual Soundness: Lessons from Static Python

### 2022-01-14 pj-7.1/paper.tex
pj: sec 5 edits

%% TODO unions tracked statically in SP, erased to dyn unless narrowed

@@

%% TODO built-in vs. user defined ... 
%% Carl: I don't think this is quite true the way it's phrased here? in general
%%   for SP classes we wrap non-static subclass override methods at runtime (in our
%%   vtable implementation) to enforce correct return type, so that we don't need
%%   exact types, we can trust that Liskov is not broken.
%%     The cases where we tend to be limited to exact types is in optimizing
%%   operations on builtin types, where they aren't really static types and
%%   methods called on them aren't going through SP vtables, so we don't have
%%   the vtable wrappers to enforce LSP, we just know the behavior of the exact
%%   builtin type

This section explains significant aspects of \SP{} that lie outside the
scope of our model.
These aspects include sound interactions with Python code,
the bytecode language and optimizations, and dynamic Python features
that \SP{} does not yet ascribe types to.

\subsection{Open-World Python Interactions}

Although \SP{} is technically a new language, it is designed for gradual adoption.
Python programmers should be able to add types one module at a time to an
existing codebase.
Consequently, \SP{} supports interactions with Python modules in the only viable
way: by letting the Python code run with minimal constraints.
After all, if programmers need to fix every module in a codebase before they can
run tests after typing a first module, then the gradual story is lost.

Interactions with open-world~\cite{vss-popl-2017} Python code pose a minor
threat to soundness.
Typed functions (and methods) cannot assume that their arguments are
well-typed.
Arguments from typed contexts can be validated by either static checks
or materialization casts, but arguments from untyped contexts are unchecked.
For this reason, \SP{} compiles every typed function to check its inputs.
\Cref{s:optimize} explains how the optimizer avoids some of these checks.
%% 2022-01-09 worth mentioning primitive types? they get cast like anything else
%% ... I guess the trick is that primitives don't materialize in typed code

FILL sound generics / checked types also need to guard their writes

There are two cases, however, where new \SP{} types do add constraints to
existing Python code.
First, sound generic types (e.g. $\sptchkdict{\sptype_0}{\sptype_1}$)
force value-suppliers to use a checked constructor.
Existing code can \emph{use}\/ a checked generic via the same API as on
un-checked one, but can create new values only using the \SP{} constructors.
Second, the strict setting for modules prevents updates to module-level variables.
If existing code happens to mutate a member of a newly-strict module, a run-time
error will occur.

@@

\SP{} can generate more-efficent code than Python because it targets
the Cinder runtime.
Cinder extends Python with bytecode instructions to check types at run-time,
to construct \SP{} data structures, and to perform optimized actions.
\Cref{t:bytecode} lists a few representative bytecode instructions.

\begin{table}
  \caption{Representative bytecode instructions}
  \label{t:bytecode}

@@

The two instructions that express run-time checks are \code{CAST}
and \code{CHECK\_ARGS}.
The former checks a value against a type.
The latter is for functions and methods; it checks all inputs to a function
against their declared types.

\Cref{t:bytecode} presents two instructions that allocate \SP{}-specific data structures.
The first creates a checked dictionary; the second creates an object with a vtable~(\cref{s:vtable}).
Cinder comes with a similar instruction to build checked lists.
The developers will add more instructions along the same lines in the future.

The remaining instructions are for optimization.
Both \bcinst{INVOKE\_FUNCTION} and \bcinst{INVOKE\_METHOD} are alternatives
to Python's dynamic call dispatch.
The former jumps to a statically-determined address;
the latter uses a v-table lookup to quickly find an address at run-time.
The load and store instructions for fields improve upon Python's generic attribute lookup.
In the JIT, these instructions can sometimes be further optimized to a single assembly instruction.
The \bcinst{FAST\_LEN} instruction gets the length of a built-in datatype directly, instead
of using the Python \code{\_\_len\_\_} attribute.
Lastly, the \bcinst{REFINE\_TYPE} instruction tells the JIT about the type of a local value whose
type is not clear from the context.

@@

Part of this goal is met by inserting casts only in positions where the dynamic
type flows into a static type.
The materialze rule in the model illustrates this policy.
Typed functions, however, are compiled with a \bcinst{CHECK\_ARGS} instruction
that casts all arguments.
By convention, this instruction always appears on the first line of a typed
function body.
The optimizer uses this convention to skip argument checks by jumping past them
when it is safe to do so; namely, whenever a typed function (or method) gets
called in a typed context.
In this way, all untyped calls get the necessary argument check and all typed
calls benefit from materialize checks at the call-site.

@@

\SP{} does not ascribe types to the following Python features.
These are not covered in our model because the implementation simply
assigns the dynamic type and lets the runtime treat them as untyped
Python code.
For each, we claim that the dynamic type is a reasonable choice;
accurate static types would be difficult to maintain.

@@

\SP{} does not attempt to type first-class classes: partly because they have
yet to appear in performance-critical code at Instagram and partly because it
is unclear how to incorporate them into the nominal type system.
The straightforward but restrictive approach is to force code that uses a
first-class class to expect subtypes of a particular named static class.
Flatt et al.~\cite{fkf-popl-1998} propose a more-flexible approach, %(specifically for mixins)
but it requires a second layer of \emph{interface types} atop the nominal hierarchy.
MonNom~\cite{mt-oopsla-2021} uses interfaces in a similar way to accommodate
structural objects.
Adding an interface layer to \SP{} is an open question.

@@

Python allows classes to inherit from a list of parents.
When resolving a method call to such a class, Python
dynamically traverses the parent list in a fixed order
seeking a first match.

Due to the dynamic method resolution order (MRO),
\SP{} does not track types for classes with multiple parents.
Method calls to these classes execute the same way as
in Python, with no type-directed optimizations to speed up dispatch.

\paragraph{Dynamic Execution}
%% TODO is eval a side channel, or safe?
%% TODO check that eval/exec args really are unoptimized / sound etc.
Results computed by calls to \code{eval} and \code{exec}
have the dynamic type.
Their inputs run without being rewritten
by the optimizer.

Studies of JavaScript and R have shown that many uses of dynamic execution
can be removed through simple adjustments~\cite{rhbv-ecoop-2011,gdkkv-oopsla-2021,mrmv-esop-2012}.
Assuming these findings carry over to Python, we recommend either similar
adjustments or the introduction of semantically-descriptive replacements
over attempting to type eval.

\paragraph{Module-Level Variables}

Any Python module can read and write to the module-level variables of another module.
When compiling a module, \SP{} therefore assumes that its module-level variables
may be modified by untyped code and assigns the dynamic type.

Cinder offers \emph{strict modules} as an alternative to the Python
semantics.\footnote{https://instagram-engineering.com/python-at-scale-strict-modules-c0bb9245c834}
If a programmer chooses to declare a module as strict, then its module-level varibles
are immutable and thus typeable.

Another potential solution is for the Cinder runtime to check that writes to module-level variables
preserve their types.
There are two downsides to this idea:
it will add some performance overhead,
and it will force developers to rewrite untyped code in order to fix any errors that arise.
%% Because of the latter concern, type dynamic is a very reasonable default.

\paragraph{Set attribute}

Carl: Removing the use of super-dynamic Python features like
\code{\_\_setattr\_\_} that we don’t support. So far in the cases where
we’ve done this it hasn’t required a major rewrite.

>>>>

@@

This section explains significant aspects of \SP{} that lie outside the
scope of our model.
These include sound interactions with Python code,
the bytecode language and optimizations, and dynamic Python features
that \SP{} does not type.

\subsection{Interactions with Open-World Python Code}

Although \SP{} is technically a new language, it is designed for gradual adoption.
Python programmers should be able to add types module-by-module to an
existing codebase.
Consequently, \SP{} supports interactions with Python modules in the only viable
way: by letting the Python code run with zero static constraints and minimal
dynamic constraints.

There are two case in which \SP{} types add dynamic constraints to Python code.
First, concrete generic types (e.g. $\sptchkdict{\sptype}{\sptype}$)
reject inputs that were not created via a checked contructor.
If a client of a Python module decides to impose concrete types, the Python module
may need to change.
Second, the strict setting for modules prevents updates to module-level variables.
If existing code mutates a binding in a strict module, a run-time error will
occur.

Interactions with \emph{open-world}~\cite{vss-popl-2017} Python code pose a minor
threat to soundness because
typed functions (and methods) cannot assume that their arguments are
well-typed.
All arguments sent from typed contexts get validated either statically or
via materialization casts, but arguments from untyped contexts are unchecked.
For this reason, \SP{} compiles every typed function to check its inputs.
Similarly, concrete types such as $\sptchkdict{\sptype}{\sptype}$ also
check their writes.
%% 2022-01-09 worth mentioning primitive types? they get cast like anything else (they don't materiazize)
%% 2022-01-13 don't think so --- leave prims alone, we don't want to frighten reviews with unsoundness

@@

\SP{} can generate efficient code because it targets
the Cinder runtime.
Cinder extends the Python 3.8 bytecode with instructions to check types at run-time,
to construct \SP{} data structures, and to perform actions in a way that benefits from
static information..

\begin{table}
  \caption{Representative Cinder bytecode instructions}
  \label{t:bytecode}

@@

\Cref{t:bytecode} lists a few representative bytecode instructions.
The two instructions that express run-time checks are \code{CAST}
and \code{CHECK\_ARGS}.
The former checks a value against a type.
The latter is for functions and methods; it checks all inputs to a function
against their declared types.

Two of the listed instructions allocate \SP{}-specific data structures.
The first creates a checked dictionary; the second creates an object with a vtable~(\cref{s:vtable}).
Cinder comes with a similar instruction to build checked lists.
%% The developers will add more instructions along the same lines in the future.

The remaining instructions are for optimization.
Both \bcinst{INVOKE\_FUNCTION} and \bcinst{INVOKE\_METHOD} are alternatives
to Python's dynamic call dispatch.
The former jumps to a statically-determined address;
the latter uses a vtable lookup to quickly find an address at run-time.
The load and store instructions for fields improve upon Python's generic attribute lookup.
In the JIT, these instructions can sometimes be further optimized to a single assembly instruction.
The \bcinst{FAST\_LEN} instruction gets the length of a built-in datatype directly, instead
of using the Python \code{\_\_len\_\_} attribute.
Lastly, the \bcinst{REFINE\_TYPE} instruction tells the JIT about the type of a local value whose
type is not clear from the context.

@@

Part of this goal is met by inserting casts only in positions where the dynamic
type flows into a static type.
The materialize rule in the model illustrates this policy.
Typed functions, however, are compiled with a \bcinst{CHECK\_ARGS} instruction
that defensively casts all arguments.
By convention, this instruction always appears on the first line of a typed
function body.
The optimizer uses this convention to skip argument checks by jumping past them
when it is safe to do so; namely, whenever a typed function or method gets
called in a typed context.

@@

\SP{} does not ascribe types to the following Python features.
These are not covered in our model because the implementation simply
assigns the dynamic type and lets the runtime treat them as untyped
Python code.
For each, the dynamic type is a very reasonable choice;
accurate static types would be burdensome to maintain.

@@

\SP{} does not attempt to type first-class classes: partly because they have
yet to appear in performance-critical code and partly because it
is unclear how to incorporate them into the nominal type system.
The straightforward but restrictive approach is to force code that uses a
first-class class to expect subtypes of a particular named static class.
Flatt et al.~\cite{fkf-popl-1998} propose a more-flexible approach, %(specifically for mixins)
but it requires a second layer of \emph{interface types} atop the nominal hierarchy.
MonNom~\cite{mt-oopsla-2021} uses interfaces in a similar way to accommodate
structural objects.
How best to add an interface layer in \SP{} is an open question.

@@

Python allows classes to inherit from a list of parents.
To resolve method calls to such a class, Python
dynamically traverses the parent list in a fixed order
seeking a first match.

Due to the dynamic method resolution order (MRO),
\SP{} does not track types for classes with multiple parents.
Method calls to these classes execute the same way as
in Python, with no type-directed optimizations to improve dispatch.

\paragraph{Dynamic Execution}

%% 2022-01-13: __setattr__ is along the same lines
%%Carl: Removing the use of super-dynamic Python features like
%%\code{\_\_setattr\_\_} that we don’t support. So far in the cases where
%%we’ve done this it hasn’t required a major rewrite.

Results computed by calls to \code{eval} and \code{exec}
have the dynamic type.
Their inputs run without being rewritten by the optimizer.

Studies of JavaScript and R have shown that many uses of dynamic execution
can be removed through simple adjustments~\cite{rhbv-ecoop-2011,gdkkv-oopsla-2021,mrmv-esop-2012}.
Assuming these findings carry over to Python, similar adjustments would
preferable over attempting to type eval.

%% 2022-01-13 covered pretty well earlier
%%\paragraph{Module-Level Variables}
%%
%%Any Python module can read and write to the module-level variables of another module.
%%When compiling a module, \SP{} therefore assumes that its module-level variables
%%may be modified by untyped code and assigns the dynamic type.
%%
%%Strict modules as an alternative to the Python
%%semantics.\footnote{https://instagram-engineering.com/python-at-scale-strict-modules-c0bb9245c834}
%%If a programmer chooses to declare a module as strict, then its module-level varibles
%%are immutable and thus typeable.
%%
%%Another potential solution is for the Cinder runtime to check that writes to module-level variables
%%preserve their types.
%%There are two downsides to this idea:
%%it will add some performance overhead,
%%and it will force developers to rewrite untyped code in order to fix any errors that arise.
%%%% Because of the latter concern, type dynamic is a very reasonable default.
